% `ss3sim` vignette
% Sean C. Anderson and lots of others...
%

Installing the `ss3sim` `R` package
===================================

The package can be installed with:

```{r install-and-load, eval=FALSE}
# dependencies if needed:
install.packages(c("r4ss", "plyr", "gtools", "ggplot2", "lubridate", "reshape2"))
install.packages("devtools")
devtools::install_github("ss3sim", "seananderson")
```

You can read the help files and access this vignette again with:

```{r help, eval=FALSE}
help(package = "ss3sim")
vignette("ss3sim")
```

`ss3sim` requires the `SS3` binary to be in your path. 
See the section "Putting `SS3` in your path" at the end of this document for details.

An example simulation with `ss3sim`
===================================

First, we'll locate three sets of folders that are located within the package
data: (1) the folder with the plaintext case files, (2) the folder with the
operating model (OM), and (3) the folder with the estimating model (EM).

```{r locate-folders}
library(ss3sim)
d <- system.file("extdata", package = "ss3sim")
case_folder <- paste0(d, "/eg-cases")
om <- paste0(d, "/models/cod-om")
em <- paste0(d, "/models/cod-em")
```

Setting up the file structure
-----------------------------

The `ss3sim` package is set up assuming that you have an established base-case OM and EM to work with. 
Each OM and EM should be in their own folder. 
The OM folder should have the files:

    yourmodel.ctl
    yourmodel.dat
    ss3.par
    starter.ss
    forecast.ss

The EM folder should have:

    yourmodel.ctl
    yourmodel.dat # optional; not used
    starter.ss
    forecast.ss

In both cases, nothing more and nothing less. 
The names of the `.ctl` and `.dat` files are not important. 
The package functions will rename them after they are copied to appropriate folders. 
These files should be derived from the `.ss_new` files but named as listed above. 
It’s important to use these `.ss_new` files so they have consistent formatting. 
Many of the functions in this package depend on that formatting.

Cases and scenarios
-------------------

To use the high-level wrapper function `run_fish600`, you will have unique case identifiers that combine to create unique scenarios. 
The types of cases are: natural mortality (M), fishing mortality (F), data quality (D), selectivity (S), growth (G), and retrospective (R). 
These case IDs are followed by three-letter species identifier.
It's important to use three letters since the functions assume that the last three letters represent a species (or some other identifier for a different project).

The different version of each case are identified with numbers. 
So, for example, the base-case scenario for a cod stock might be: `D0-E0-F0-G0-M0-R0-S0-cod`. 
The order of the letters doesn't matter, as long as you use them consistently.


Output file structure
---------------------

The function `copy_ss3models` creates a folder structure and copies over
the operating and estimation models. The folder structure looks like:

      D0-E0-F0-G0-M0-R0-S0-cod/1/om
      D0-E0-F0-G0-M0-R0-S0-cod/1/em
      D0-E0-F0-G0-M0-R0-S0-cod/2/om
      D0-E0-F0-G0-M0-R0-S0-cod/2/em
      ...

If you are using bias adjustment (`bias_adjust = TRUE`) then there will
be some additional folders. In that case the folders will look like:

      D0-E0-F0-G0-M0-R0-S0-cod/bias/1/om
      D0-E0-F0-G0-M0-R0-S0-cod/bias/1/em
      D0-E0-F0-G0-M0-R0-S0-cod/bias/2/om
      D0-E0-F0-G0-M0-R0-S0-cod/bias/2/em
      ...
      D0-E0-F0-G0-M0-R0-S0-cod/1/om
      D0-E0-F0-G0-M0-R0-S0-cod/1/em
      D0-E0-F0-G0-M0-R0-S0-cod/2/om
      D0-E0-F0-G0-M0-R0-S0-cod/2/em
      ...

Note that the operating and estimating model folders have been renamed `om` and `em` within each iteration, 
the operating and estimation models have been checked to make sure they contain the minimal files (as listed above), 
the filenames have been made all lowercase, 
the data file has been renamed `data.dat`, 
the control files have been renamed `om.ctl` or `em.ctl`, 
and the starter and control files have been adjusted to reflect these new file names.

The functions in this package assume you've set your working directory in `R` to be the base folder where you will store the scenario folders.

Creating the input configuration files
--------------------------------------

You will need to have a folder containing "case" argument definitions. 
These plain text files are read by `get_caseval` and turned into argument lists that are passed to `run_ss3sim`. 
You can create template input files by running `create_argfiles`. 
It reads the various functions and parses the arguments and default values into plain text files. 
The default settings create these files:

1.  `M0-spp.txt`

2.  `F0-spp.txt`

3.  `index0-spp.txt`

4.  `agecomp0-spp.txt`

5.  `lcomp0-spp.txt`

6.  `R0-spp.txt`

7.  `S0-spp.txt`

8.  `G0-spp.txt`

9.  `E0-spp.txt`

Look in your working directory for the template files. 
Change the case ID number (defaults to `0`) and the species identifier to a three letter identifier. 
For the FISH 600 project use one of `cod`, `sar`, or `fla` for cod, sardine, or flatfish. 
An example filename would be `M1-sar.txt` or `lcomp2-fla.txt`. 
The case `D1` corresponds to the files `index1-spp.txt`, `agecomp1-spp.txt`, and `lcomp0-spp.txt`. 
The other case types have single argument files.

The first column in the text files denotes the argument to be passed to a function. 
The second argument denotes the value to be passed. 
You can use any simple `R` syntax. 
For example: `c(1, 2, 4)`, or `seq(1, 100)` or `1:100` or `matrix()`. 
Character objects don’t need to be quoted, but can be if you’d like. 
However, be careful not to use the delimiter (set up as a semicolon) anywhere else in the file besides to denote columns. 
You can add comments after any `#` symbols. 
Internally, the functions evaluate in `R` any entries that have no character values (e.g. 
`1:100`) or have an alpha-numeric character followed by a `(`. 
Anything that is character only or has character mixed with numeric but doesn't have the regular expression `"[A-Za-z0-9]("` gets turned into a character
argument.

Putting that all together, here’s what an example `F1-cod.txt` file might look like:

    years; 1:100
    years_alter; NA 
    fvals; NA

Model checking with deterministic simulations
---------------------------------------------

We'll run some "deterministic" runs to check our model for bias when we don't have any process error.
To do this, we'll start by setting up a matrix of recruitment deviations with 0 deviations. 
We need 100 rows (for 100 year simulations) and 20 columns (for 20 deterministic iterations).

```{r }
recdevs_det <- matrix(0, nrow = 100, ncol = 20)
```

Then we'll set up case "estimation" files in which the recruitment deviations are set to the nominal level of `0.001`. 
We'll name these files `E100-cod.txt` and `E101-cod.txt`. 
In the control files, the key element is setting `par_name = SR_sigmaR` and `par_int = 0.001`.

When we run the simulations, we'll pass our deterministic recruitment deviations to the function `run_fish600`. 
Running 20 replicates should be enough to identify whether our models are performing as we expect.

```{r deterministic-runs, eval=FALSE}
run_fish600(iterations = 1:20, scenarios = 
  c("D0-E100-F0-G0-R0-S0-M0-cod",
    "D1-E100-F0-G0-R0-S0-M0-cod", 
    "D0-E101-F0-G0-R0-S0-M0-cod",
    "D1-E101-F0-G0-R0-S0-M0-cod"), 
  case_folder = case_folder, om_model_dir = om, em_model_dir = em,
  bias_adjust = TRUE, user_recdevs = recdevs_det)
```

Running the stochastic simulations
----------------------------------

Now we can run the stochastic simulations.

```{r stochastic-runs, eval=FALSE}
run_fish600(iterations = 1:100, scenarios = 
  c("D0-E0-F0-G0-R0-S0-M0-cod",
    "D1-E0-F0-G0-R0-S0-M0-cod", 
    "D0-E1-F0-G0-R0-S0-M0-cod",
    "D1-E1-F0-G0-R0-S0-M0-cod"), 
  case_folder = case_folder, om_model_dir = om, em_model_dir = em,
  bias_adjust = TRUE)
```

The function `get_results_all` reads in a set of scenarios and combines the output into two `.csv` files: `final_results_scalar.csv` and `final_results_ts.csv`.

```{r get-results, eval=FALSE}
get_results_all(user.scenarios = 
  c("D0-E100-F0-G0-R0-S0-M0-cod"),
    "D1-E100-F0-G0-R0-S0-M0-cod", 
    "D0-E101-F0-G0-R0-S0-M0-cod",
    "D1-E101-F0-G0-R0-S0-M0-cod", 
    "D0-E0-F0-G0-R0-S0-M0-cod",
    "D1-E0-F0-G0-R0-S0-M0-cod",
    "D0-E1-F0-G0-R0-S0-M0-cod",
    "D1-E1-F0-G0-R0-S0-M0-cod"))
```

Let's read in the `.csv` files:

```{r read-output, eval=FALSE}
scalar_dat <- read.csv("final_results_scalar.csv")
ts_dat <- read.csv("final_results_ts.csv")
```

```{r load-output, echo=FALSE}
# this is to save file space in the package:
load("ts_dat.rda")
load("scalar_dat.rda")
```

And calculate some useful values in new columns:

```{r transform-output}
scalar_dat <- transform(scalar_dat,
  SSB_MSY=(SSB_MSY_em-SSB_MSY_om)/SSB_MSY_om,
  log_max_grad = log(max_grad))

scalar_dat <- transform(scalar_dat, 
  steep = (SR_BH_steep_om - SR_BH_steep_em)/SR_BH_steep_om, 
  logR0 = (SR_LN_R0_om - SR_LN_R0_em)/SR_LN_R0_om, 
  depletion = (depletion_om - depletion_em)/depletion_om,
  SSB_MSY = (SSB_MSY_em - SSB_MSY_om)/SSB_MSY_om, 
  SR_sigmaR = (SR_sigmaR_em - SR_sigmaR_om)/SR_sigmaR_om, 
  NatM = (NatM_p_1_Fem_GP_1_em - NatM_p_1_Fem_GP_1_om)/NatM_p_1_Fem_GP_1_om)

ts_dat <- transform(ts_dat, SpawnBio = (SpawnBio_em - SpawnBio_om)/SpawnBio_om)
ts_dat <- merge(ts_dat, scalar_dat[,c("scenario", "replicate",
    "max_grad")])

scalar_dat_det <- subset(scalar_dat, E %in% c("E100", "E101"))
scalar_dat_sto <- subset(scalar_dat, E %in% c("E0", "E1"))
ts_dat_det <- subset(ts_dat, E %in% c("E100", "E101"))
ts_dat_sto <- subset(ts_dat, E %in% c("E0", "E1") & replicate %in% 1:50)
```

Now let's look at boxplots of the deterministic model runs. 

```{r plot-deterministic, fig.height=3.5, fig.width=9, fig.cap="Boxplot of relative error for SSB MSY. We see relatively little bias."}

plot_scalar_boxplot(scalar_dat_det, x = "SR_LN_R0_om", y = "SSB_MSY",
  vert = "D", vert2 = "E", relative_error = TRUE)

# add more here
```

Let's look at the relative error in estimates of spawning biomass. 
We'll colour the time series according to the 

```{r plot-sto-ts, fig.height=3.5, fig.width=9, fig.cap="Time series of relative error in spawning stock biomass."}
plot_ts_points(ts_dat_sto, y = "SpawnBio", vert = "D", vert2 = "E", 
  color = "max_grad", relative_error = TRUE)
```

```{r test-plots, fig.height=5, fig.width=5, fig.cap="Test plots"}
library(ggplot2)
p <- ggplot(ts_dat_sto, aes(year, SpawnBio_em, group = replicate)) + geom_line(alpha = 0.2) + facet_grid(D~E)
print(p)

```

Putting `SS3` in your path
==========================

`SS3` must be in your path for the `ss3sim` package to work. Your “path”
is a list of folders that your operating system looks in whenever you
type the name of a program on the command line. Having a binary in your
path means that your operating system knows where to look for the file
regardless of what folder you’re working in.

For Unix (Linux and OS X)
-------------------------

To check if `SS3` is in your path: open a Terminal window and type
`which SS3` and hit enter. If you get nothing returned then SS is not in
your path. The easiest way to fix this is to move the `SS3` binary to a
folder that’s already in your path. To find existing path folders type
`echo $PATH` in the terminal and hit enter. Now move the `SS3` binary to
one of these folders. For example, in a Terminal window type:

      sudo cp ~/Downloads/SS3 /usr/bin/

You will need to use `sudo` and enter your password after to have
permission to move a file to a folder like `/usr/bin/`.

If you’ve previously modified your path to add a non-standard location
for the `SS3` binary, you may need to also tell `R` about the new path.
The path that `R` sees may not include additional paths that you've
added through a configuration file like `.bash_profile`. You can add to
the path that `R` sees by including a line like this in your `.Rprofile`
file. (This is an invisible file in your home directory.)

    Sys.setenv(PATH=paste(Sys.getenv("PATH"),"/my/non-standard/folder",sep=":")) 

For Windows
-----------

To check if SS is in your path: open a DOS prompt and type `ss3 -?` and
hit enter. If you get a line like “ss3 is not recognized ...” then SS is
not in your path. To add it to your path:

1.  Find the latest version of the `ss3.exe` binary on your computer

2.  Record the folder location. E.g. `C:/SS3.24o/`

3.  Click on the start menu and type “environment”

4.  Choose “Edit environment variables for your account” under Control
    Panel

5.  Click on `PATH` if it exists, create it if doesn't exist

6.  Choose `PATH` and click edit

7.  In the “Edit User Variable” window add to the **end** of the
    “Variable value” section a semicolon and the `SS3` folder location
    you recorded earlier. E.g. `;C:/SS3.24o/`

8.  Restart your computer

9.  Go back to the DOS prompt and try typing `ss3 -?` and hitting return
    again.
