\documentclass[letterpaper,11pt]{letter}
\usepackage{geometry}
\geometry{letterpaper}
\usepackage{graphicx}
\usepackage{fontspec}
\usepackage{pdfsync}
\defaultfontfeatures{Ligatures=TeX}
\setsansfont{Calibri}
\setmonofont{Inconsolata}
\setmainfont{Linux Libertine}
\topmargin -2cm
\textheight 22.5cm
\address{Sean C. Anderson\\Biological Sciences\\
  Simon Fraser University\\Burnaby, BC, V5A 1S6\\
  Canada\\sean@seananderson.ca}

\begin{document}
\begin{letter}{}
\pagestyle{empty}
\opening{Dear Editor}

% You should supply an approximately one page cover letter that:
% Concisely summarizes why your paper is a valuable addition to the scientific literature
% Briefly relates your study to previously published work
% Specifies the type of article you are submitting (for example, research article, systematic review, meta-analysis, clinical trial)
% Suggests appropriate PLOS ONE Academic Editors to handle your manuscript (view a complete listing of our academic editors)
% Lists any recommended or opposed reviewers

Please find enclosed a manuscript entitled ``ss3sim: An R package for fisheries
stock assessment simulation with Stock Synthesis'' that we wish to submit for
possible publication as a research article in PLOS ONE.

Simulation testing is an important component to developing, verifying, and
understanding increasingly complex fishery stock assessment methods. One of the
more widely used stock assessment methods is Stock Synthesis 3 (SS3). However,
there lacks a generalized framework for simulation testing with SS3 --- most
stock assessment simulation-testing work to date has used custom frameworks
tailored to the particular needs of each study.

In our paper, we introduce ss3sim, an R package that facilitates large-scale,
rapid, and reproducible simulation testing. We describe how ss3sim works,
illustrate a simple example with code, compare the package to other related
software, and outline important research questions our package could address.
We also include an R vignette as Supplementary Material, which describes in
greater detail how to use the package and contains fully reproducible example
code. To our knowledge, this is the first generalized and open-source software
package focused specifically on fisheries stock assessment stimulation testing.

We confirm that this manuscript has not been published elsewhere and is not
under consideration by another journal. All authors have approved the
manuscript and agree with its submission to PLOS ONE.

We would like to suggest the following possible independent referees:

\begin{description}
  \item [Mark N. Maunder] mmaunder@iattc.org, Inter-American Tropical Tuna
    Commission, 8604 La Jolla Shores Drive, La Jolla, CA, 92037-1508, USA
  \item [Kevin R. Piner] kevin.piner@noaa.gov, NOAA Fisheries, SWFSC, 8604 La
    Jolla Shore Drive, La Jolla, CA 92037, USA
  \item [Arni Magnusson] arnima@hafro.is, Marine Research Institute, Skulagata
    4, PO Box 1390, 121 Reykjavik, Iceland
  \item [Alan C. Hicks] EMAIL ADDRESS
  \item [Jonathan J. Deroba] jonathan.deroba@noaa.gov, National Marine Fisheries
    Service, Northeast Fisheries Science Center, 166 Water Street, Woods Hole,
    MA 02543, USA
  \item [Simon D. Hoyle] EMAIL Oceanic Fisheries Program, Secretariat of the Pacific
    Community, Nouméa CEDEX, New Caledonia
  \item [David B. Sampson] david.sampson@oregonstate.edu, Oregon State
    University, Coastal Oregon Marine Experiment Station, 2030 S. Marine Science
    Drive Newport, OR, 97365, USA
\end{description}

Thank you for considering our manuscript for publication and we look forward to
hearing from you.

\bigskip
\closing{Sincerely,\\
Sean Anderson, Cole Monnahan,\\Kelli Johnson, Kotaro Ono,\\Juan Valero}

\end{letter}
\end{document}
